Aneuploidy can be an evolutionary detour \\ on the path to adaptation

### 2022-10-02 ms/ms.tex
updated manuscript
target size discussion

% without back transitions:
The estimated beneficial mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.025\ [1.024-1.026]$,
$w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
Thus, we can infer that the cost of chromosome III trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of the beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).

@@

% parameter estimates
\paragraph{Rate and fitness effect of aneuploidy and mutation.}
We inferred the rates of aneuploidy and mutation and their effects on fitness. 
We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, higher than a previous estimate of $6.7 \cdot 10^{-6}$~\citep{Zhu2016}. This may be due to genetic instability caused by heat stress~\citep{Chen2012a}.
In addition, we find no evidence for increased mutation rates in aneuploid cells. 
Previous empirical studies have suggested that genetic instability (e.g., elevated mutation rates) in aneuploid cells is due to stress associated with the aneuploid state~\citep{Bouchonville2009, Chen2012b, Zhu2012, Ippolito2021b}.
However, in the experiment of \citet{Yona2012}, both the wild-type and the aneuploid were under heat stress, which may explain why we did not find evidence for an increased mutation rate.

% TODO caveats - MP

@@

% FIG MAP frequency dynamics in log scale
\begin{figure}[p]
    \centering
	\includegraphics[width=\textwidth]{../figures/MAP_freq_dynamics.pdf}
	\caption{\textbf{Posterior predicted genotype frequencies in log-log scale.}
	Frequency dynamics of the different genotypes with MAP parameter estimates, same as \Cref{fig:FA}A, but in log-log scale. Black dashed curves for a deterministic model without genetic drift. 
	Clearly, appearance of \anwt\ and \eumtM\ is deterministic.
	Appearance of \anmt, and therefore \eumtA, is stochastic, however, the frequency dynamics are deterministic above a frequency of roughly 0.001. 
	}
    \label{fig:MAP_freq_dynamics_log}    
\end{figure}

>>>>

% without back transitions:
The estimated beneficial mutation rate is $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$. From the literature, the mutation rate per base pair is roughly $2-3\cdot10^{-10}$ (refs.~\citep{Zhu2014,Lynch2008}), but it may be higher under heat stress, as several stresses may cause hypermutation in yeast~\citep{Heidenreich2007}.
If we assume a 10-fold increase in mutation rate, then the estimated beneficial mutation rate can be explained by a target size of 1,000 base pairs.
\citet{Yona2012} found at least 10 genes on chromosome III with mutations that led to increased heat tolerance. Assuming that other chromosomes also have a similar number of heat-tolerance genes (i.e., chromosome III is one of the smaller chromosomes~\citep{Gilchrist2019}), and that each gene has a target size of ten base pairs, we easily get to a target size above 1,600.

The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval). 
However, this difference may be explained by an increased aneuploidy rate during heat stress~\citep{Chen2012a}.

The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.025\ [1.024-1.026]$,
$w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
Thus, we can infer that the cost of chromosome III trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of the beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).

@@

% parameter estimates
\paragraph{Rate and fitness effect of aneuploidy and mutation.}
We inferred the rates of aneuploidy and mutation and their effects on fitness. 
We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, higher than a previous estimate of $6.7 \cdot 10^{-6}$~\citep{Zhu2016}. This may be due to genetic instability caused by heat stress~\citep{Chen2012a}.
In addition, we find no evidence for increased mutation rates in aneuploid cells. 
Previous empirical studies have suggested that genetic instability (e.g., elevated mutation rates) in aneuploid cells is due to stress associated with the aneuploid state~\citep{Bouchonville2009, Chen2012b, Zhu2012, Ippolito2021b}.
However, in the experiment of \citet{Yona2012}, both the wild-type and the aneuploid were under heat stress, which may explain why we did not find evidence for an increased mutation rate specifically in aneuploid cells.

@@

% FIG MAP frequency dynamics in log scale
\begin{figure}[p]
    \centering
	\includegraphics[width=\textwidth]{../figures/MAP_freq_dynamics.pdf}
	\caption{\textbf{Posterior predicted genotype frequencies in log-log scale.}
	Frequency dynamics of the different genotypes with MAP parameter estimates, same as \Cref{fig:FA}A, but in log-log scale. Black dashed curves for a deterministic model without genetic drift. 
	Clearly, appearance of \anwt\ and \eumtM\ is deterministic.
	Appearance of \anmt, and therefore \eumtA, is stochastic, however, the frequency dynamics are deterministic above a frequency of roughly 0.001. 
	Note that the \eumtM\ and the \eumtA+\eumtM\ lines are overlapping for much of their trajectories.
	}
    \label{fig:MAP_freq_dynamics_log}    
\end{figure}
